---
title: "Week 9: Data Import with readr and readxl"
---

## Learning Objectives

By the end of this chapter, you should be able to:

- Import CSV and TSV files with `readr`
- Read Excel files with `readxl`
- Understand how column types are parsed
- Parse dates, times, and numbers correctly
- Diagnose and fix import problems

---

# 1. Reading CSV and TSV Files

The `readr` package (part of the tidyverse) provides fast and friendly functions for reading text data.

### Example: Reading a CSV file

```{r}
library(tidyverse)

df <- read_csv("https://people.sc.fsu.edu/~jburkardt/data/csv/airtravel.csv")
glimpse(df)
```

### Example: Reading a TSV file

```{r, eval=FALSE}
df_tsv <- read_tsv("data/example.tsv")
```

---

## In-Class Exercise 1 – CSV/TSV

1. Download a small CSV file (e.g., from the course repository).  
2. Read it into R using `read_csv()`.  
3. Inspect its structure with `glimpse()` and `summary()`.  
4. What data types were automatically detected?

---

# 2. Column Types and Parsing

`readr` automatically guesses column types, but you can override them.

### Example: Overriding column types

```{r, eval=FALSE}
df <- read_csv("data/mydata.csv", col_types = cols(
  id = col_character(),
  date = col_date(format = "%Y-%m-%d")
))
```

You can parse numbers with `parse_number()`, dates with `parse_date()`, and times with `parse_time()`.

---

## In-Class Exercise 2 – Parsing

1. Create a vector of messy numbers: `c("$100", "250%", "300")`.  
2. Use `parse_number()` to extract numeric values.  
3. Create a vector of dates as strings and use `parse_date()`.

---

# 3. Importing Excel Files

The `readxl` package is used to read Excel files (`.xls`, `.xlsx`).

### Example: Reading an Excel sheet

```{r, eval=FALSE}
library(readxl)

excel_df <- read_excel("data/example.xlsx", sheet = "Sheet1")
head(excel_df)
```

---

## In-Class Exercise 3 – Excel Import

1. Use a provided Excel file (or download one).  
2. Read the first sheet with `read_excel()`.  
3. Specify a different sheet and check the result.  

---

# 4. Handling Import Problems

When column parsing fails:

- Use `problems()` to diagnose
- Use `col_types` to fix column types
- Clean data after import using `mutate()`

Example:

```{r, eval=FALSE}
bad <- read_csv("data/bad.csv")
problems(bad)
```

---

# 5. Reading Other Formats (Optional)

- `read_delim()` – for custom delimiters
- `read_table()` – for whitespace-delimited files
- `read_lines()` – for line-by-line text
- `jsonlite::fromJSON()` – for JSON files (optional preview)

---

## In-Class Challenge – Import & Clean Workflow

1. Import a messy CSV file with mixed types.  
2. Fix incorrect column parsing.  
3. Convert a date column to proper Date format.  
4. Summarize the data by a grouping variable.

---

# Homework Preview

For Homework 6, you will:

- Import at least **one CSV** and **one Excel** dataset
- Fix any parsing issues (e.g., column types, dates)
- Clean at least one column with `mutate()`
- Provide a short summary (using `group_by()` and `summarize()`)
- Render to PDF and submit

---

# Next Steps

Next week, we will learn to **work with text data and regular expressions** using the `stringr` package.
