---
title: "Relational Data with dplyr Joins"
---

## Learning Objectives

By the end of this chapter, you should be able to:

-   Understand the concept of relational data and keys
-   Combine multiple datasets using different join functions
-   Use `left_join()`, `inner_join()`, `full_join()`, and `semi_join()`
-   Diagnose and handle join problems (missing keys, duplicates)
-   Apply joins in analysis workflows

------------------------------------------------------------------------

## What is Relational Data?

Relational data consists of **multiple tables** that can be linked by **keys**.

Example tables from `nycflights13`:

-   `flights`: flight information
-   `airlines`: airline names
-   `airports`: airport locations
-   `planes`: plane details
-   `weather`: weather data

------------------------------------------------------------------------

## Keys

-   **Primary key**: uniquely identifies each row in a table\
-   **Foreign key**: column that matches a primary key in another table

Example: `flights$carrier` matches `airlines$carrier`.

------------------------------------------------------------------------

## Joins with dplyr

`dplyr` join functions merge tables by keys.

### `left_join()`

Keeps all rows from the first table:

```{r}
library(tidyverse)
library(nycflights13)

flights |>
  left_join(airlines, by = "carrier") |>
  rename(airline_name = name) |>
  select(airline_name, carrier, flight) |>
  head()
```

------------------------------------------------------------------------

### `inner_join()`

Keeps only matching rows:

```{r}
flights |>
  inner_join(airlines, by = "carrier") |>
  rename(airline_name = name) |>
  select(airline_name, carrier, flight) |>
  head()
```

------------------------------------------------------------------------

### `full_join()`

Keeps all rows from both tables:

```{r}
flights |>
  full_join(airlines, by = "carrier") |>
  rename(airline_name = name) |>
  select(airline_name, carrier, flight) |>
  head()
```

------------------------------------------------------------------------

### `semi_join()` and `anti_join()`

-   `semi_join()`: keeps rows in first table with matches in second\
-   `anti_join()`: keeps rows with no matches

```{r}
flights |>
  semi_join(airlines, by = "carrier") |>
  head()
```

------------------------------------------------------------------------

### In-Class Exercise 1 – Basic Joins

1.  Use `left_join()` to add airline names to `flights` (rename to `airline_name`).\
2.  Count the number of flights for each airline.\
3.  Use `inner_join()` and compare the number of rows.

------------------------------------------------------------------------

## Joining Multiple Tables

You can chain joins to combine several datasets:

```{r}
flights |>
  left_join(airlines, by = "carrier") |>
  rename(airline_name = name) |>
  left_join(airports, by = c("dest" = "faa")) |>
  select(airline_name, dest, arr_delay) |>
  head()
```

------------------------------------------------------------------------

### In-Class Exercise 2 – Multi-Table Joins

1.  Join `flights` with `airports` to add destination airport names.\
2.  Summarize average arrival delay by airport.\
3.  Which airport has the longest average delay?

------------------------------------------------------------------------

## Handling Join Problems

-   Missing keys → results in `NA` values\
-   Duplicated keys → may create duplicate rows\
-   Always check results with `count()` or `distinct()`

Example:

```{r}
flights |>
  left_join(airlines, by = "carrier") |>
  rename(airline_name = name) |>
  count(carrier, airline_name)
```

------------------------------------------------------------------------

## In-Class Challenge – Join Workflow

-   Join flights with airlines and airports\
-   Calculate average arrival delay by airline and destination\
-   Arrange by delay and identify the worst-performing routes

------------------------------------------------------------------------

## Homework Preview

For the next homework, you will:

-   Combine at least two datasets using joins
-   Use at least two different join types (`left_join()`, `inner_join()`, etc.)
-   Handle missing data or duplicates appropriately
-   Produce a summary table and one visualization based on the joined data
-   Render to PDF and submit on Canvas

------------------------------------------------------------------------

## Next Steps

Next, we will introduce **accessing data** using spreadsheets, SQL databases, JSON, and web scraping.
