---
title: "Introduction to R and RStudio"
---

## Learning Objectives

By the end of this chapter, you should be able to:

-   Install and open R and RStudio\
-   Navigate the four-pane layout of RStudio\
-   Create and run R scripts\
-   Understand the differences between the console, script editor, and environment\
-   Execute basic R operations and understand data types\
-   Install and load R packages\
    \## Getting Started

**R** is a programming language designed for data analysis.\
**RStudio** is an Integrated Development Environment (IDE) that makes working with R easier.

### Installing R and RStudio

1.  Install R: <https://cran.r-project.org/>\
2.  Install RStudio: <https://posit.co/download/rstudio-desktop/>

Once you open RStudio, you'll see four panes:

-   **Console** – for running code interactively\
-   **Source** – to write and save scripts\
-   **Environment/History** – to view and manage objects\
-   **Files/Plots/Packages/Help/Viewer** – navigation and visualization tools

------------------------------------------------------------------------

## Basic R Concepts

### Variables and Assignments

```{r}
x <- 5
y <- 10
z <- x + y
z
```

### Vectors and Functions

```{r}
ages <- c(25, 30, 35, 40)
mean(ages)
sd(ages)
```

### Data Frames

```{r}
name <- c("Alice", "Bob", "Charlie")
age <- c(25, 30, 35)
student_data <- data.frame(name, age)
student_data
```

### Inspecting Data

```{r}
str(student_data)
summary(student_data)
head(student_data)
```

### Comments and Help

```{r}
# This is a comment
?mean  # Help for the mean function
```

### Using Scripts and Console

-   Write your code in the script editor and run lines with Ctrl+Enter (Cmd+Enter on Mac)
-   Save your work with .R extensions
-   Use the Console for quick exploration

### Installing and Loading Packages

```{r install, eval=FALSE}
# Run this in the Console if you haven't installed yet
# install.packages("tidyverse")
```

## Practice Exercises

1.  Create a script that assigns two numbers and prints their sum
2.  Create a vector of Celsius temperatures and convert to Fahrenheit
3.  Create a data frame with columns: name, age, major

```{r}
# Exercise 1
a <- 15
b <- 20
a + b

# Exercise 2
celsius <- c(0, 10, 20)
fahrenheit <- celsius * 9/5 + 32
fahrenheit

# Exercise 3
students <- data.frame(
  name = c("Lily", "Mark", "Tom"),
  age = c(21, 22, 23),
  major = c("Biology", "Math", "History")
)
str(students)
```

## More In-Class Exercises

-   Create a vector of 5 numbers and calculate the mean, median, and standard deviation.
-   Create a data frame with three columns (e.g., name, age, and major) and print its structure.
-   Import a dataset using `read.csv()` from a URL and summarize it using `summary()`.

```{r}
my_vec <- c(10, 20, 30, 40, 50)
mean(my_vec)
median(my_vec)
sd(my_vec)

df <- data.frame(
  name = c("Lily", "Mark", "Tom"),
  age = c(21, 22, 23),
  major = c("Biology", "Math", "History")
)
str(df)

data <- read.csv("https://people.sc.fsu.edu/~jburkardt/data/csv/airtravel.csv")
summary(data)
```

## Homework Preview

-   Write a script to create three different vectors and compute descriptive statistics.
-   Import a CSV dataset of your choice, explore it with `str()` and `summary()`, and write one paragraph about what you've learned.

## Next Steps

Now that you can run R scripts and explore objects, you'll learn how to load and reshape real data using the `tidyverse` in the next chapter.
