---
title: "Strings and Regular Expressions with stringr"
---

## Learning Objectives

By the end of this chapter, you should be able to:

- Manipulate strings using the `stringr` package
- Detect patterns with regular expressions (regex)
- Extract, replace, and split text
- Clean messy text data for analysis

---

## Introduction to stringr

The `stringr` package provides consistent, simple functions for string operations.

Load the library:

```{r}
library(stringr)
```

---

## Creating and Inspecting Strings

```{r}
fruit <- c("apple", "banana", "pear")
str_length(fruit)
str_c(fruit, " is tasty")
```

---

### In-Class Exercise 1 – Basic String Operations

1. Create a vector of at least 5 words.  
2. Measure their lengths with `str_length()`.  
3. Concatenate them with the phrase `" is cool"`.

---

## Detecting Patterns with Regex

`str_detect()` returns TRUE if a pattern is found.

```{r}
words <- c("dog", "cat", "parrot", "cow")
str_detect(words, "o")
```

You can use **regular expressions** for more complex patterns.

Examples:

- `^a` – starts with “a”  
- `ing$` – ends with “ing”  
- `[0-9]+` – one or more digits

```{r}
animals <- c("ant", "bat", "cat", "dog")
str_detect(animals, "^a")
```

---

### In-Class Exercise 2 – Pattern Detection

1. Create a vector of email-like strings.  
2. Use `str_detect()` to check which contain `"@"`.  
3. Write a regex to detect strings ending in `.com`.

---

## Extracting and Replacing Text

### `str_extract()`

Extracts the first match:

```{r}
str_extract(c("abc123", "xyz789"), "[0-9]+")
```

### `str_replace()`

Replaces matching patterns:

```{r}
str_replace("apple pie", "apple", "peach")
```

---

### In-Class Exercise 3 – Extraction and Replacement

1. Extract digits from a vector of alphanumeric strings.  
2. Replace the word `"dog"` with `"puppy"` in a text vector.

---

## Splitting and Cleaning Text

### `str_split()`

Splits text into pieces:

```{r}
str_split("a,b,c", ",")
```

### Cleaning with regex

You can remove unwanted characters:

```{r}
dirty <- c(" price:$100 ", " cost:$200 ")
str_replace_all(dirty, "[$ ]", "")
```

---

### In-Class Challenge – Text Cleaning

1. Create a vector of messy product names with extra spaces and symbols.  
2. Use `str_replace_all()` and `str_trim()` to clean them.  
3. Extract numeric prices from the strings.

---

## Homework Preview

For the next homework, you will:

- Work with a text dataset (e.g., movie titles, email logs, or messy product names)
- Use at least three `stringr` functions to clean or extract information
- Write one regex pattern to detect a specific feature in the data
- Render a short report (with code and results) to PDF and submit

---

## Next Steps

Next, we will learn to **work with factors and categorical data** using the `forcats` package.
