---
title: "Tidy Data with tidyr"
---

## Learning Objectives

By the end of this chapter, you should be able to:

-   Explain why tidy data improves analysis and visualization
-   Reshape data between wide and long formats using `pivot_longer()` and `pivot_wider()`
-   Separate and unite columns using `separate()` and `unite()`
-   Apply tidying techniques to messy real-world datasets
-   Prepare datasets for use with `dplyr` and `ggplot2`

------------------------------------------------------------------------

## Why Tidy Data?

In Week 6, you performed **EDA** on datasets that were already in a usable format.\
Real datasets are often messy. **Tidy data** makes it easy to:

-   Use `ggplot2` for visualization\
-   Use `dplyr` for summaries and transformations\
-   Combine datasets with joins

**Principles of Tidy Data** (Hadley Wickham):\
1. Each variable is a column\
2. Each observation is a row\
3. Each value is a cell

------------------------------------------------------------------------

## Pivoting: Long vs Wide

### `pivot_longer()`

Converts wide data into long (tidy) format.

```{r}
library(tidyverse)

table4a |>
  pivot_longer(cols = c(`1999`, `2000`),
               names_to = "year",
               values_to = "cases")
```

------------------------------------------------------------------------

### `pivot_wider()`

Converts long data back into wide format.

```{r}
table2 |>
  pivot_wider(names_from = type, values_from = count)
```

------------------------------------------------------------------------

### In-Class Exercise 1 – Pivoting

1.  Use `pivot_longer()` to convert `table4a` to long format.\
2.  Use `pivot_wider()` on `table2` to create separate columns for `type`.\
3.  Which format is easier to use with `ggplot2` and `dplyr`?

------------------------------------------------------------------------

## Separating and Uniting Columns

### `separate()`

Splits a column into multiple columns.

```{r}
table3 |>
  separate(rate, into = c("cases", "population"), sep = "/")
```

------------------------------------------------------------------------

### `unite()`

Combines multiple columns into one.

```{r}
table5 |>
  unite(new, century, year, sep = "")
```

------------------------------------------------------------------------

### In-Class Exercise 2 – Separate and Unite

1.  Use `separate()` to split the `rate` column in `table3`.\
2.  Use `unite()` to combine `century` and `year` into one column.

------------------------------------------------------------------------

## Tidying a Real Dataset

The `who` dataset is messy: column names encode multiple variables.

Example tidying workflow:

```{r}
who |>
  pivot_longer(cols = starts_with("new"),
               names_to = "key",
               values_to = "cases",
               values_drop_na = TRUE) |>
  separate(key, into = c("type", "sex_age"), sep = "_") |>
  separate(sex_age, into = c("sex", "age"), sep = 1)
```

------------------------------------------------------------------------

### In-Class Exercise 3 – WHO Dataset

1.  Pivot `who` longer to create `key` and `cases`.\
2.  Separate `key` into multiple components.\
3.  Count total cases by country.\
4.  Which country has the highest reported cases?

------------------------------------------------------------------------

## Tidy Data Workflow

After tidying, you can:

-   Use `ggplot2` for visualizations\
-   Use `group_by()` and `summarize()` for summaries\
-   Join with other datasets

------------------------------------------------------------------------

## Homework Preview

For homework, you will:

-   Take a messy dataset (e.g., `table4a`, `table5`, or your own)
-   Use `pivot_longer()` and/or `pivot_wider()` to reshape it
-   Use `separate()` and `unite()` as needed
-   Produce a tidy dataset and create **one visualization** and **one grouped summary**
-   Render to PDF and submit on Canvas

------------------------------------------------------------------------

